--- 
title: "Spatial Social Networks (SSN) Visualization and Metrics with R"
author: "Xiaofan Liang, Clio Andris, Dipto Sarkar"
date: "2023-05-01"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This book describes the basic concepts, theories, methods, and visualizations involved to map and analyze spatial networks in R."
---

# Introduction 

This tutorial is dedicated to describe the basic concepts, methods, and visualizations involved to map and analyze non-planar networks (social networks & relationships) in geographic space using R. A non-planar network is a network in which the nodes are geolocated, while the edges can cross over each other without creating new nodes. For example, mapping Facebook friendships and flight routes in geographic space are non-planar. A planar network is spatially embedded in a way that the edges intersect only at the endpoints, such as the road, river, and eletrical grid networks. There are already [some tutorials](https://www.r-spatial.org/r/2019/09/26/spatial-networks.html) for mapping and analyzing planar networks, but none exists for non-planar networks. 

We assume our audience has basic knowledge of R programming with `tidyverse` syntax and basic GIS concepts (e.g., basic geometry types, coordinate system projection, spatial operation etc.). If not, you can use this book [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-operations.html) to get started with the basic concepts. The book also leverages `tmap` at map visualization. 

Mapping relationship in geographic space has its own challenges, which is described thoroughly in Andris's paper [Challenges for Social Flows](https://isiarticles.com/bundles/Article/pre/pdf/105132.pdf). In our visualization examples, we will try to show the best practices to map social networks in space that can counter some of these problems. 

All the data you need to complete the tutorial can be loaded from `SSNtools` R package, which is developed together with this tutorial.

**Required Packages**

We use `igraph` package to calculate network metrics, `sf` package to create geospatial objects from dataframe, `tmap` package to create both static and interactive maps, `basemaps` package to extract OpenStreetMap basemaps, and `tidyverse` package as the syntax to process data. You can download the packages using the following codes:

```{r eval=FALSE}
install.packages("igraph") #for network analysis
install.packages("sf") #for spatial objects operation
install.packages("tmap") #for interactive and static maps 
install.packages("basemaps") #for retrieving basemaps for tmap 
install.packages("tidyverse") #for data processing 
install.packages('tigris') #for loading TIGER shapefiles
install.packages('stplanr') #for using od2line function to convert points to lines
install.packages('ggplot2') #for visualizing statistical graphics
install.packages("devtools") #to download R package on GitHub
devtools::install_github("friendlycities-gatech/SSNtools") #install development version of our SSNtools R package for advanced SSN metrics. 
```

There are a few packages for visualizing maps. We chose `tmap` in the end for its flexibility to switch between interactive maps and static plots. We also have some codes that use `ggplot2` to generate other graphic visualizations. 

`ggplot2` has powerful mapping functions. It is intuitive for people who are used to the underlying logic of ggplot2. One advantage of `ggplot2` is that it can easily add high-resolution Google Maps as basemaps using `ggmap` package (only in WGS84 projection). Here are some good tutorials that teach you how to use ggplot2 to [map planar networks](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html) and [produce basic elements of a map](https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/make-maps-with-ggplot-in-R/). However, you need to convert spatial objects to dataframe for `ggpplot2` to draw on.  

`tmap` is equally powerful as `ggpplot2`, and has the extra function to switch between a static plot and interactive plot mode. This feature is useful when you want to zoom in and out to explore your data.  

`sf` converts data frame into spatial objects that can be used in base R plot function and `tmap`. It can also conduct various spatial operation, such as coordinate system projection, spatial joins, and topological relations. A brief introduction of how to streamline `sf` and `tmap` is given in the book [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-operations.html). 

`igraph` is the go-to package for network analysis, such as calculating the centrality measures of a network. These measures may become the attributes for nodes or edges in the maps (see [Chapter 3](https://friendlycities-gatech.github.io/SSN_tutorial/nodes.html#nodes)). 

`tidyverse` provides a set of packages that share the same data representation and use the [pip symbol](https://style.tidyverse.org/pipes.html) `%>%` to connect steps of processing. 

`tigris` download geographic boundary shapefiles from TIGER website.

`stplanr` is an R package that analyze planar spatial network. We mostly use the `od2line()` function in the package to convert points to lines. 

`SSNtools` is an R package that provides advanced metrics for analyzing and visualizing spatial social networks. It also include all the sample data we use in the tutorial. 

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'igraph', 'sf', 'tmap', 'tidyverse', 'basemaps', 'stplanr', 'devtools', 'SSNtools'
), 'packages.bib')
```
